### 2024-10-20 Article/UniversalConstructions.tex
reordering + redefn of graphs eqgrph

\section{Limits and Universal Constructions}\label{sect:limits_univ_constr}

\todo{aggiungere i diagrammi e sistemare per l'indice}

@@

% ---- Coni, Limiti ----

@@

\begin{remark}
    It makes sense to refer to a (co)limit as \emph{the} (co)limit. Suppose $(P, p_i)_{i\in \cat I}$ and $(Q, q_i)_{i \in \cat I}$ be limits for a diagram $D: \cat{I \rightarrow C}$. Then, there exists a unique morphism $h: Q \rightarrow P$ such that $p_i \circ k = q_i$ for each $i$. At the same way, there exists a unique morphisms $k: P \rightarrow Q$ such that $q_i \circ k = p_i$ for each $i$. From the existence of the identity, must be $k \circ h = id_{Q}$ and $h \circ k = id_P$, that is, $P$ and $Q$ are isomorphic.
\end{remark}

Notion such limits and colimits are generalization of more particular cases that will be now introduced, that we will often call \emph{universal constructions}.

% ---- Oggetti Iniziali & Terminali ----

\begin{definition}[Initial Object, Terminal Object]
    Consider the empty diagram (i.e., a diagram $D: \textbf{0} \rightarrow \cat C$ where $\textbf{0}$ is the empty category \Cref{ex:0_cat}). Then, the limit of $D$ is called \emph{terminal object} and the colimit of $D$ is called \emph{initial object}, denoted, respectively, $\terminal_\cat{C}$ and $\initial_{\cat C}$. (Subscripts are omitted when they are clear from the context).
\end{definition}

\begin{example}\label{ex:set_init_term}
    In $\Set$, the initial object is the empty set $\varnothing$, because, for each set $S$, there exists the empty function from $\varnothing$ to $S$. The terminal object of $\Set$ is the singleton $\{ \bullet \}$, because there is exactly one function from a set $S$ to $\{ \bullet \}$, namely, the function which sends each $s \in S$ to $\bullet$.
    % It is possible to visualize the \Cref{obs:terminal_are_isomorph}: given two singletons $\{ \bullet \}$ and $\{ \star \}$, the function between them is bijective, while there exists a unique initial object.
\end{example}

% ----- Prodotti, Coprodotti ----

\begin{definition}[Product, Coproduct]
        Let $D$ be the following diagram:
    \[
        \begin{tikzcd}
            A & & B
        \end{tikzcd}
    \]
    Then, a cone for $D$ is an object $X$ and two arrows $f: X \rightarrow A$, $g: X \rightarrow B$ (i.e., a span, defined in \Cref{ex: span}):
    \[
        \begin{tikzcd}
            A & X \arrow[l, "f" swap] \arrow[r, "g"] & B
        \end{tikzcd}
    \]
    If it exists, a limit for $D$ is called \emph{product} of $A$ and $B$, usually denoted as $(A\times B, \pi_A, \pi_B)$, while whose arrows are called \emph{projections}.
    The colimit of $D$ is called \emph{coproduct} of $A$ and $B$, usually denoted as $(\iota_A, \iota_B, A + B)$.
\end{definition}

The notions of product and coproduct can be easily generalized, extending the definition to the product (and coproduct) of a family of objects, together with appropriate arrows (e.g., the projection arrows for each object in the product). We will denote the product of a collection of objects indexed by a (finite) category $\cat I$ as $\big(\prod_{i \in \Ob(\cat I)} X_i, (\pi_i)_{i \in \Ob(\cat I)}\big)$, and the coproduct as \- $\big((\iota_i)_{i \in \Ob(\cat I)}, \coprod_{i \in \Ob(\cat I)} X_i\big)$.

% ---- Equalizzatori, Coequalizzatori ----

\begin{definition}[Equalizer, Coequalizer]\label{def:equalizer_coequalizer}
    Let $D$ be the diagram below.
    \[
        \begin{tikzcd}
            A \arrow[r, shift left, "f"] \arrow[r, shift right,"g"swap] & B
        \end{tikzcd}
    \]
    The limit of $D$ is called \emph{equalizer}, and its colimit is called \emph{coequalizer}.
\end{definition}

\begin{proof}
    Suppose $X$ be an object and $x, y: X \rightarrow E$ be two morphisms in $\cat C$ such that $e \circ x = e \circ y$, and let $z = e \circ x$. Then, since $e$ is an equalizer, $f \circ e = g \circ e$, and $f \circ z = g \circ z$. But, for the universal property of limits, there must be exactly one $u: Z \rightarrow E$ such that $z = e \circ u$. It follow that $x = u$ and $y = u$, hence $x = y$.
\end{proof}

% ---- Pullback, Pushout -----

\begin{definition}[Pullback, Pushout]\label{def:pullback_pushout}
    Let $D$ be the cospan $(f, C, g) : A \rightharpoondown B$. % $$(A \xrightarrow{f} C \xleftarrow{g} B)$.
     A cone for $D$ is an object $P$ and three arrows $\phi:P \rightarrow A$, $\psi: P \rightarrow B$, and $h: P \rightarrow C$, but the latter is uniquely determined by the other ones ($f \circ \phi = h = g \circ \psi$).
    Thus, the following diagram is a cone:
    \[
        \begin{tikzcd}
            P \arrow[r, "{\psi}"] \arrow[d, "{\phi}" swap] & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
     Then, the limit of $D$ is called \emph{pullback} of $f$ and $g$.
     Given a span $S = (l, X, r): L \rightharpoonup R$, shown in the diagram below,
    \[
        \begin{tikzcd}
            L & X \arrow[l, "l"swap] \arrow[r, "r"]& R
        \end{tikzcd}
    \]
    a cocone for $S$ is any commutative square of the form
    \[
        \begin{tikzcd}
            & C & \\
            L \arrow[ur, "f"] &
            X \arrow[l, "l"swap] \arrow[r, "r"]&
            R \arrow[ul, "g" swap]
        \end{tikzcd}
    \]
    (the morphism $X \rightarrow C$ is uniquely determined by the relation $f \circ l = g \circ r$).
    The colimit for $S$ is called \emph{pushout} of $l$ and $r$.
\end{definition}

The proposition above can be dualised stating that pushouts preserves epimorphisms.

>>>>

\section{Universal Constructions}\label{sect:univ_constr}
\todo{Questa è da riscriviere e ridurre}
% What we are interested in are the "best" construction having a certain property, and we exprtess this fact via requiring unique factorization when another construction does

\subsection{Initial and Terminal Objects}

The next definitions are about \emph{universal constructions}. The simplest ones are the notion of initial and, dually, terminal objects.

\begin{definition}[Initial and terminal object]
    An object $A$ of a category $\cat C$ is said to be \emph{initial} if, for each other object $B$ of $\cat C$, there exists a unique morphism from $A$ to $B$.
    Dually, an object $Z$ is said to be a terminal object in a category $\cat C$ if, for any other object $X$ of $\cat C$, there exists a unique morphism from $X$ to $Z$.
    An initial object is indicated by the symbol $\initial$, and a terminal object by the symbol $\terminal$.
\end{definition}

\begin{obs}\label{obs:terminal_are_isomorph}
    It makes sense to refer to an initial (and terminal) object as \emph{the} initial (\emph{the} terminal) object. Suppose that $\initial$ and $\initial'$ are distinct initial objects of a category $\cat C$. Then, there exists a unique morphism from $\initial$ to $\initial'$, say $f$. Likewise, it must exist a unique morphism from $\initial'$ to $\initial$, say $g$. Then, $g\circ f$ must be exactly $id_{\initial}$, and $f \circ g = id_{\initial'}$, and then they are isomorphic. The same argument works for the terminal object.
\end{obs}

\begin{example}\label{ex:set_init_term}
    In $\Set$, the initial object is the empty set $\varnothing$, because, for each set $S$, there exists the empty function from $\varnothing$ to $S$. The terminal object of $\Set$ is the singleton $\{ \bullet \}$, because there is exactly one function from a set $S$ to $\{ \bullet \}$, namely, the function which sends each $s \in S$ to $\bullet$. It is possible to visualize the \Cref{obs:terminal_are_isomorph}: given two singletons $\{ \bullet \}$ and $\{ \star \}$, the function between them is bijective, while there exists a unique initial object.
\end{example}

\subsection{Products and Coproducts}

More complex constructions are products (and, dually, coproducts).

\begin{definition}[Product]
    A \emph{product} of two objects $A$ an $B$ is an object $A \times B$ together with two \emph{projection arrows} $\pi_1:A\times B \rightarrow A$ and $\pi_2: A \times B \rightarrow B$ such that, for every object $C$ and pair of arrows $f: C \rightarrow A$, $g: C \rightarrow B$, there is exactly one arrow $\langle f, g \rangle : c \rightarrow A \times B$ making the diagram commutes
    \[
        \begin{tikzcd}
                    & C \arrow[dl, "f" swap] \arrow[d, dashed, "{\langle f, g \rangle}" description] \arrow[dr, "g"] & \\
                A   & A \times B  \arrow[l, "{\pi_1}"] \arrow[r, "{\pi_2}" swap]    &  B    
        \end{tikzcd}
    \]
   -- i.e., such that $\pi_1 \circ \langle f, g \rangle = f$ and $\pi_2 \circ \langle f, g \rangle  = g$.

\begin{definition}[Coproduct]
    The dual of the product is the \emph{coproduct}.
    A coproduct of two objects $A$ and $B$ is an object $A + B$ together with two arrows $\iota_1:A\rightarrow A+B$, $\iota_2: B \rightarrow A + B$ such that, for every object $C$ and pair of arrows $f:A \rightarrow C$, $g:B \rightarrow C$, there is a unique arrow $[f, g] : A + B \rightarrow C$ such that the diagram commutes
    \[
        \begin{tikzcd}
            A \arrow[r, "{\iota_1}"] \arrow[dr, "f" swap] & A+B \arrow[d, dashed, "{[f,g]}" description] & B \arrow[l, "{\iota_2}" swap] \arrow[dl, "g"] \\
            & C & 
        \end{tikzcd}
    \]
    -- i.e., such that $[f, g] \circ \iota_1 = f$ and $[f, g] \circ \iota_2 = g$.
\end{definition}

The notions of product and coproduct can be easily generalized, extending the definition to the product (and coproduct) of a family of objects, together with appropriate arrows (e.g., the projection arrows for each object in the product). We will denote the product of a collection of objects indexed by a (finite) category $\cat I$ as $\big(\prod_{i \in \Ob(\cat I)} X_i, (\pi_i)_{i \in \Ob(\cat I)}\big)$, and the coproduct as $\big((\iota_i)_{i \in \Ob(\cat I)}, \coprod_{i \in \Ob(\cat I)} X_i\big)$.

Again, the definition of these constructions is divided into two parts: one stating what the construction is, and another stating that the construction satisfies the universal property.

\subsection{Equalizers and Coequalizers}

The next notion is about a construction that make parallel arrows (i.e., two possibly distinct morphisms with the same domain and codomain) equal. In $\Set$, given two functions $f, g: A \rightarrow B$, such construction corresponds exactly to the restriction of the domain to the set of elements on which $f$ and $g$ are equal. Specifically, if we take the set $E = \{x \in A \mid f(x) = g(x)\}$, and we take the function $e: E \rightarrow A$ defined by $e(x) = x$, we obtain what is called an \emph{equalizer} for $f$ and $g$. In fact, $e$ is that function that make $f$ and $g$ be the same function -- i.e., $f \circ e = g \circ e$.
This concept can be generalized as follows.

\begin{definition}[Equalizer and Coequalizer]
    Let $f, g: A \rightarrow B$ be two arrows of a category $\cat C$. An \emph{equalizer} for $f$ and $g$ is pair $(E, e)$, where $E$ is an object and $e: E \rightarrow A$ is an arrow in $\cat C$ such that:
    \begin{enumerate}
        \item \label{eq_1} $ f \circ e = g \circ e$; and
        \item if $(E', e')$ is another pair that satisfies~\ref{eq_1}, then there exists a unique $h: E' \rightarrow E$ such that $e \circ h = e'$.
    \end{enumerate}
    \[
        \begin{tikzcd}
            E \arrow[r, "e"] & A \arrow[r, shift left, "f"] \arrow[r, shift right, "g" swap] & B \\
            E' \arrow[u, dashed, "k"] \arrow[ur, "e'" swap] & 
        \end{tikzcd}
    \]

    A \emph{coequalizer} of $f$ and $g$, dually, is a pair $(c, C)$, where $C$ is an object and $c: B \rightarrow C$ such that $c \circ f = c \circ g$, with the universal property.
\end{definition}

\begin{proof}
    Suppose $X$ be an object and $x, y: X \rightarrow E$ be two morphisms in $\cat C$ such that $e \circ x = e \circ y$, and let $z = e \circ x$. Then, since $e$ is an equalizer, $f \circ e = g \circ e$, and $f \circ z = g \circ z$. But, for the universal property of equalizers, there must be exactly one $u: Z \rightarrow E$ such that $z = e \circ u$. It follow that $x = u$ and $y = u$, hence $x = y$.
\end{proof}

\subsection{Pullbacks and Pushouts}

Given two arrows, another pair of constructions one can find in a category is given by pullbacks and pushouts.

\begin{definition}[Pullback, Pushout]
        A \emph{pullback} of two arrows $f: A \rightarrow C$ and $g: B \rightarrow C$ is a triple $(P, \pi_g, \pi_f)$ containing an object $P$ and a pair of arrows $\pi_g: P \rightarrow A$, $\pi_f: P \rightarrow B$ such that:
        \begin{enumerate}
            \item\label{pb_1} $f \circ \pi_g = g \circ \pi_f$; and
            \item if $(X, p, q)$, where $p: X \rightarrow A$ and $q: X \rightarrow B$, satisfies~\ref{pb_1}, then there is a unique $k:X \rightarrow P$ such that $p = \pi_g \circ k$ and $j = \pi_f \circ k$.
        \end{enumerate}
        \[
        \begin{tikzcd}
        X \arrow[drr, bend left=20, "q"] \arrow[ddr, bend right=20,"p" swap] \arrow[dr, dashed, "k"] & & \\
        & P  \arrow[d, "{\pi_g}" swap] \arrow[r, "{\pi_f}"] & B \arrow[d, "g"] \\
        & A  \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]

    Dually, the \emph{pushout} of two arrows $f: A \rightarrow B$ and $g: A\rightarrow C$ is a triple $(\iota_g, \iota_f, I)$, where $I$ is an object and $\iota_g:B \rightarrow I$ and $\iota_f: C \rightarrow I$ are morphisms such that:
    \begin{enumerate}
        \item\label{po_1} $\iota_g \circ f = \iota_f \circ g$; and 
        \item if $(i, j, Y)$, where $i: B \rightarrow Y$ and $j: C \rightarrow Y$, satisfies~\ref{po_1}, then there is a unique $h: I \rightarrow Y$ such that $i = h \circ \iota_g$ and $j = h \circ \iota_f$.  
    \end{enumerate}
    \[\begin{tikzcd}
        A & C \\
        B & I \\
        && Y
        \arrow["g", from=1-1, to=1-2]
        \arrow["f" swap, from=1-1, to=2-1]
        \arrow["{{\iota_f}}", from=1-2, to=2-2]
        \arrow["j", bend left=20, from=1-2, to=3-3]
        \arrow["{{\iota_g}}" swap, from=2-1, to=2-2]
        \arrow["i" swap, bend right=20, from=2-1, to=3-3]
        \arrow["h", dashed, from=2-2, to=3-3]
    \end{tikzcd}\]
    {% \[
        %     \begin{tikzcd}
        %         Y & &\\
        %         & I \arrow[ul, dashed, "h"] & B \arrow[l, "{\iota_g}" swap] \arrow[ull, bend right=20, "i"swap] \\
        %         & C \arrow[u, "{\iota_f}"] \arrow[uul, bend left=20, "j"] & A \arrow[l, "g"] \arrow[u, "f" swap]
        %     \end{tikzcd}
        % \]
    }
\end{definition}

Pullbacks (and, dually, pushouts) are a construction that is slightly more general than products and equalizers. An intuition of what they represent is given by considering what is concretely a pullback in the category of sets.

@@

We now introduce the notion that generalize all the universal constructions defined above. In fact, initial objects, products, equalizers and pullbacks (dually, terminal objects, coproducts, coequalizers and pushouts) can be seen as the particular case of a certain type of construction, called limit.

To define what a limit is, we first need to define cones.

@@

The following examples show how limits are general concepts for the constructions defined above.

\begin{example}
    Let $D$ be the empty diagram in the category $\cat C$. Then, a cone for $D$ is any object of $\cat C$, while a limit is the terminal object of $\cat C$.
\end{example}

\begin{example}\label{ex:product_are_limits}
    Let $D$ be the following diagram:
    \[
        \begin{tikzcd}
            A & & B
        \end{tikzcd}
    \]
    Then, a cone for $D$ is an object $X$ and two arrows $f: X \rightarrow A$, $g: X \rightarrow B$ (i.e., a span, defined in \Cref{ex: span}):
    \[
        \begin{tikzcd}
            A & X \arrow[l, "f" swap] \arrow[r, "g"] & B
        \end{tikzcd}
    \]
    If it exists, a limit for $D$ is the product of $A$ and $B$.
\end{example}

\begin{example}\label{ex:pb_as_limit}
    A pullback is the limit for the diagram below.
    \[
        \begin{tikzcd}
            & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
    In fact, a cone for the diagram above is an object $P$ and three arrows $\phi:P \rightarrow A$, $\psi: P \rightarrow B$, and $h: P \rightarrow C$, but the latter is uniquely determined by the other ones ($f \circ \phi = h = g \circ \psi$).
    Thus, the following diagram is a cone:
    \[
        \begin{tikzcd}
            P \arrow[r, "{\psi}"] \arrow[d, "{\phi}" swap] & B \arrow[d, "g"] \\
            A \arrow[r, "f" swap] & C
        \end{tikzcd}
    \]
    For $(P, \phi, \psi)$ to be a pullback, it must have the universal property. In other words, it has to be a limit.

    This example show us that a pullback is a limit for a cospan (\Cref{ex: span}).
\end{example}

\begin{example}\label{ex:equaliz_are_limits}
    A limit for the diagram below is an equalizer for $f$ and $g$.
    \[
        \begin{tikzcd}
            A \arrow[r, shift left, "f"] \arrow[r, shift right,"g"swap] & B
        \end{tikzcd}
    \]
\end{example}

All the examples we provided about limits are still valid, in their dual form, for colimits. In fact, initial objects, coequalizers, coproducts and pullbacks are particular cases of limits, and the way to see this fact is the same seen in previous examples.

\begin{example}
    Given a span $S = (l, X, r): L \rightharpoonup R$, shown in the diagram below,
    \[
        \begin{tikzcd}
            L & X \arrow[l, "l"swap] \arrow[r, "r"]& R
        \end{tikzcd}
    \]
    a cocone for $S$ is any commutative square of the form
    \[
        \begin{tikzcd}
            & C & \\
            L \arrow[ur, "f"] &
            X \arrow[l, "l"swap] \arrow[r, "r"]&
            R \arrow[ul, "g" swap]
        \end{tikzcd}
    \]
    (the morphism $X \rightarrow C$ is uniquely determined by the relation $f \circ l = g \circ r$).
    A colimit for $S$ is then a pushout of $l$ and $r$.
\end{example}
